---
title: "MH O-Ring Example"
author: "Jonathan Navarrete"
date: "September 6, 2017"
output: 
  html_document:
    css: slides.css
---





# Bayesian Analysis

## O-ring Challenger Data { .selectable }


The following is a well covered logistic regression example using the O-ring data set related to the 1986 space shuttle Challenger exploision. The output is modeled as the probability of failure (Y = 1) given the data.

$$
P(Y = 1 | X = x) = p = \frac{exp(\alpha + x \beta)}{1 + exp(\alpha + x \beta)}
$$
Or, equivalently with the logit transformation on $P$
$$
logit(p) = \frac{p}{1 - p} = \alpha + x \beta
$$



## O-ring Challenger Data { .selectable }

```{r}

failure <- c(1, 1, 1, 1, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 
             0, 0, 0, 1, 0, 0, 0, 0, 0)
temperature <- c(53, 57, 58, 63, 66, 67, 67, 67, 68, 
                 69, 70, 70, 70, 70, 72, 73, 75, 75,
                 76, 76, 78, 79, 81)

df = data.frame(failure, temperature)
head(df)

```



## O-ring Challenger Data { .selectable }

The frequentist logistic regression

```{r, echo = TRUE}
fit = glm(formula = failure ~ temperature, data = df, 
          family = binomial(link = "logit"))

summary(fit)

```



## O-ring Challenger Data { .selectable }


Observed are 
$$
Y_i \sim Bernoulli(p(x_i)), \text{ where } p(x_i) = \frac{exp(\alpha + x_i \beta)}{1 + exp(\alpha + x_i \beta)}
$$
where $p(x_i)$ is the probability of O-ring failure at temperature $x_i$. The likelihood is
$$
L(\alpha, \beta | \mathbf{y}) \propto \prod^n_{i = 1} \left(\frac{exp(\alpha + x_i \beta)}{1 + exp(\alpha + x_i \beta)} \right)^{y_i} \times \left(\frac{ 1 }{1 + exp(\alpha + x_i \beta)} \right)^{1 - y_i}
$$

and as a prior Robert & Casella choose
$$
\pi_{\alpha}(\alpha | b) \times \pi_{\beta}(\beta) = \frac{1}{b} e^{\alpha} e^{- e^{\alpha} / b}
$$
which puts an exponential prior on $log(\alpha)$ and a flat prior on $\beta$ (uniform), and insures a proper posterior distribution. Note that priors on $\alpha$ and $\beta$ are independent. This will be important for computational purposes in the M-H algorithm.

The prior above is an exponential distribution with $Exp(\lambda)$, where $\lambda =  \frac{1}{b} e^{\alpha}$


## O-ring Challenger Data { .selectable }

```{r}
## Output from ML estimation from Logistic Regression
## MLEs make great starting valules
a.mle <- as.numeric(fit$coefficients[1])
b.mle <- as.numeric(fit$coefficients[2])
var.a.mle <- summary(fit)$cov.scaled[1, 1]
var.b.mle <- summary(fit)$cov.scaled[2, 2]

b.hyper <- exp(a.mle + 0.577216) ## hyper parameter
## 0.577216 is "Euler’s constant"


```




## O-ring Challenger Data { .selectable }


Let's set up the posterior and proposal distributions

```{r}
## setting up functions

# Posterior distribution
dPosterior <- function(theta, y = failure, x = temperature){
    ## density of Y is binomial/bernoulli
    a <- theta[1]
    b <- theta[2]
    p <- 1 - 1 / (1 + exp(a + b * x)) ## logistic CDF
    lik <- exp(sum(dbinom(y, size=1, prob=p, log=TRUE)))
    dprior <- exp(a) * exp(-exp(a) / b.hyper) * 1/b.hyper ## density of prior
    return(lik * dprior)
}
```



## O-ring Challenger Data { .selectable }

```{r}
# Proposal distribution (independent proposal, so "theta0" is not used)
dProposal <- function(theta){
    ## ignore theta0
    a <- theta[1]
    b <- theta[2]
    # a <- log(rexp(1, 1 / b.hyper)) ## remember, log for computational purposes
    # try: exp(rexp(1, 1/b.hyper)) ## Inf
    pr1 <- exp(a) * exp(-exp(a) / b.hyper) * 1/b.hyper
    pr2 <- dnorm(b, b.mle, sqrt(var.b.mle))
    return(pr1 * pr2)
}
```


## O-ring Challenger Data { .selectable }

```{r}
rProposal <- function(theta0){
    ## independent proposals for a and b
    #a <- log(rexp(1, 1 / b.hyper))
    a <- log(rexp(1, 1 / b.hyper)) ## log for computational purposes
    b <- rnorm(1, b.mle, sqrt(var.b.mle))
    return(c(a, b))
}

```


## O-ring Challenger Data { .selectable }

Now we run the M-H algorithm

```{r}
## Metropolis-Hastings set up to run
# Run Metropolis-Hastings
N = 1000000
BurnIn =  5000

```


## O-ring Challenger Data { .selectable }

```{r}
## x0: initializing values for MH algorithm
## f: the pdf of posterior; target distribution
## dProposal: the pdf of the proposal distribution
## rProposal: the random number gen for proposal
## N: Number of samples needed
## BurnIn: Number of burn in samples
## function generates N+B samples then drops B
x0 <- c(a.mle, b.mle)
x <- matrix(NA, nrow = N + BurnIn, ncol = length(x0)) ## empty matrix
fx <- rep(NA, N + BurnIn)
x[1,] <- x0 ## add the initializing values
fx[1] <- dPosterior(x0) ## initiate the M-H algorithm
accept <- 0 ## counter for acceptance

for(i in 2:(N + BurnIn)){
    u <- rProposal(x[i-1,])
    fu <- dPosterior(u)
    ## use log values for computational purposes
    r <- log(fu) + log(dProposal(x[i-1,])) - log(fx[i-1]) - log(dProposal(u))
    Rho <- min(exp(r), 1)
    if(runif(1) <= Rho){
        accept <- accept + 1 ## update counter
        x[i,] <- u
        fx[i] <- fu
    } else {
    x[i,] <- x[i-1,]
    fx[i] <- fx[i-1]
    }
}
```


## O-ring Challenger Data { .selectable }

```{r}
print("acceptance rate: ")
print(accept/(N+BurnIn))

MH.Results <- x[-(1:BurnIn), ]
alphaMH <- MH.Results[,1]
betaMH <- MH.Results[,2]


```




## O-ring Challenger Data | Summary { .selectable }


```{r}

summary(MH.Results) ## summary of parameters
```


## O-ring Challenger Data { .selectable }

```{r}
par(mfrow = c(1,2))
hist(alphaMH, freq=FALSE, col="gray", border="white", xlab=expression(alpha), main = "")
plot(alphaMH, type="l", col="gray", xlab="Iteration", ylab=expression(alpha), main = "")
lines(1:N, cumsum(alphaMH) / (1:N))
```


## O-ring Challenger Data { .selectable }

```{r}
par(mfrow = c(1,2))
hist(betaMH, freq=FALSE, col="gray", border="white", xlab=expression(beta), main = "")
plot(betaMH, type="l", col="gray", xlab="Iteration", ylab=expression(beta), main = "")
lines(1:N, cumsum(betaMH)/(1:N))

```

